---
params: 
    pTRIAL: "ALL"
title: 'Longitudinal `r params$pTRIAL`'
#author: "Adrian Cantu (Adapted from Christopher Nobles)"
date: "`r format(Sys.time(), '%Y %B %d')`"
output: 
  pdf_document:
#    latex_engine: lualatex
    toc: false
#    keep_md: true
#    keep_tex: true
    number_sections: false
    toc_depth: 2
    fig_caption: true
    df_print: default
    highlight: espresso
header-includes: 
  - \usepackage{float,indentfirst,booktabs,longtable,array,multirow,pdflscape,tabu}
  - \usepackage[normalem]{ulem}
  - \setlength{\defaultaddspace}{0em}
  - \setlength{\parindent}{2em}
fontsize: 11pt
geometry: margin=0.75in
knit: (
  function(inputFile, encoding) { 
  
    ptrial <- 'CALL'
  
    rmarkdown::render( 
      input       = inputFile, 
      encoding    = encoding, 
      params      = list(pTRIAL = ptrial),      
      output_file = paste(ptrial,'_logitudinal_', Sys.Date(), '.pdf', sep='')) })
---

```{r setup, include=FALSE}
library(tidyverse)
library(kableExtra)

if(!exists('workingDir')) {workingDir <- "/home/ubuntu/data/CART19/CART19_from_git2"}
if(!exists('utilsDir')) {utilsDir <- "/home/ubuntu/data/CART19/CART19_from_git2/utils"}

knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center",
  comment = "",
  echo = FALSE,
  warning = FALSE,
  error = TRUE,
  message = FALSE,
  cache = FALSE,
  dpi = 300,
  dev = "png",
  results = "asis",
  fig.pos = "H",
  fig.width = 7,
  dev.args = list(type = "cairo-png")
)
TRIAL <- params$pTRIAL
nnn<-500 # max cutoff to show in the plot (will only affect drowing, not results)
p_b <- 0.2 # proportion of total breaks used to define cutoff
```


```{r}
# Import HGNC reference data for annotation and consistency ----
hgnc_complete <- data.table::fread(
  paste0("zcat ", file.path(utilsDir, "hgnc_complete_set.180207.txt.gz")),
  sep = "\t", header = TRUE, 
  select = c(
    "HGNC ID", "Approved Symbol", "Approved Name", "Locus Group", 
    "Locus Type", "Synonyms", "Previous Symbols", "Entrez Gene ID", 
    "Ensembl Gene ID", "RefSeq (supplied by NCBI)", 
    "UniProt ID (supplied by UniProt)"
  ),
  data.table = FALSE
)

names(hgnc_complete) <- c(
  "hgnc_id", "symbol", "name", "locus_group", "locus_type", "alias_symbol", 
  "prev_symbol", "entrez_id", "ensembl_gene_id", "refseq_accession", 
  "uniprot_ids"
)

hgnc_complete <- dplyr::filter(hgnc_complete, !grepl("withdrawn", symbol)) %>%
  dplyr::mutate(
    kegg_id = paste0("hsa:", entrez_id),
    entrez_id = paste0(entrez_id, ":EZID"),
    alias_symbol = gsub(", ", "|", alias_symbol),
    prev_symbol = gsub(", ", "|", prev_symbol),
    extended_alias = paste0(
      alias_symbol, "|", prev_symbol, "|", ensembl_gene_id, "|", 
      refseq_accession, "|", uniprot_ids))

```

```{r}
# redoo relevant parts of gene impact
# gene_impact %>% select(TP_TDN_max_span,max_span,max_time,TP_TDN_max_time)
# df3 %>% select(TP_TDN_max_span,max_span,TP_num_patients,TDN_num_patients,TP_TDN_num_patients)
# stats_tp_tdn %>% filter(TP_TDN_first_time!=TP_TDN_last_time)
#df3
```


```{r}

gene_impact_tmp <- readRDS( # TODO, not sure where this comes from
  file.path(workingDir,paste0('ONLY_',TRIAL),
            paste0("only_",TRIAL,"_cart19_gene_impact.rds"))) 


gene_impact <- gene_impact_tmp %>%
  dplyr::select(
    -ort_fisher_test, -ort_fisher_test_CR, -ort_fisher_test_NR
  ) %>%
  #dplyr::filter(!gene_name %in% excluded_gene_names) %>%
  dplyr::mutate(gene_name = spraphal::alias_arbiter(
    IDs = gene_name,
    RefIDs = hgnc_complete$symbol,
    aliasIDs = hgnc_complete$extended_alias,
    sep = "|", remove_absent_IDs = NULL, quiet = TRUE))


total_breaks_v <- gene_impact %>%
    dplyr::filter(
      TP_TDN_num_patients >= 3,
      TP_TDN_num_sites >= 3,
    ) %>%
  pull(TP_TDN_sum_abund

       ) %>% 
  sort(.,decreasing = TRUE)
total_breaks <-   sum(total_breaks_v)

first_top <- which(cumsum(total_breaks_v) > total_breaks*p_b)[1]

break_num_gene <- tibble(x=c(1:nnn),
  y= sapply(c(1:nnn), function(xx){
  gene_impact %>%
    dplyr::filter(
      TP_TDN_num_patients >= 3,
      TP_TDN_num_sites >= 3,
      TP_TDN_sum_abund >= xx,
    ) %>%
  nrow() 
}))

break_rank_gene <- tibble(x=c(1:length(total_breaks_v)),
  y= total_breaks_v)


df3 <- gene_impact %>%
  dplyr::filter(
    TP_TDN_num_patients >= 3,
    TP_TDN_num_sites >= 3,
    TP_TDN_sum_abund >= total_breaks_v[first_top],
#    TP_TDN_max_span>5
  )

df3.2 <- df3 %>%
  dplyr::arrange(
    desc(TP_TDN_time_top3avg),desc(TP_TDN_max_span), desc(max_time), desc(TP_TDN_long_count), desc(TP_TDN_num_sites),
    desc(TP_TDN_num_patients), On_Onco_List
  ) %>%
  dplyr::select(
    gene_name, TP_TDN_time_top3avg,TP_TDN_max_span, max_time, TP_TDN_long_count, TP_TDN_num_patients, TP_TDN_num_sites,
    TP_TDN_peak_abund,TP_TDN_peak_relAbund ,On_Onco_List
  ) %>%
  dplyr::mutate(TP_TDN_peak_relAbund=paste0(as.character(round(TP_TDN_peak_relAbund,digits = 2)),'%'),
      TP_TDN_time_top3avg=round(TP_TDN_time_top3avg,digits = 2) ) %>% 
  dplyr::rename(
    "Gene" = gene_name, "Num. Timepoints" = TP_TDN_long_count, "Time Span" = TP_TDN_max_span,
    "Longest Time" = max_time, "Patients" = TP_TDN_num_patients,
    "Sites" = TP_TDN_num_sites, "Max Abund." = TP_TDN_peak_abund,
    "Max RelAbund" = TP_TDN_peak_relAbund,
    "Onco-Related" = On_Onco_List, "t3avg"=TP_TDN_time_top3avg
  )

```

Insertion sites that are close to Genes (less than 5,000bp) are grouped and labeled with that
gene name. Longitudinal gene of interest are those that have insertion sites in at least 3 different positions in at least 3 patients. Additionally, a cutoff of at least `r total_breaks_v[first_top]` independent sonic breaks is imposed. This number is chosen such that the selected genes represent at lease `r  paste0(as.character(p_b*100),'%')` of all sonic breaks (see plot)  
The table has the following columns:

* **gene**: Gene name according to the HUGO Gene nomenclature committee (version from 2018-02-07)
* **t3avg**: average of last time point for the 3 latest insertion sites.
* **Time Span**: Maximum time between the first and last sample that have the same insertion in this gene and originate from the same patient.
* **Longest time**: Last time point(in days) where insertion are observed.
* **Num. Timepoints**: Number of times points the most common insertion site is observed.
* **Patients**: Number of patient with at least one insertion near this gene.
* **Sites**: Number of distinct insertion positions.
* **Max Abund**: Maximum sonic abundance observed for an insertion site near this gene.
* **Max RelAbund**: Maximum relative abundance observed for an insertion site near this gene.
* **Onco-Related**: Whether the gene is in the cancer related list


```{r fig1, fig.cap="Effect of filtering by sonic breaks", out.height='40%'}
ggplot(data=break_num_gene, aes(x=x, y=y)) +
  geom_line() +
  theme_classic() +
  labs(x="Required sonic breaks",
       y = "Number of genes") +
  geom_vline(xintercept = total_breaks_v[first_top],linetype = "dashed") +
  geom_text(aes(total_breaks_v[first_top]-nnn*0.01,length(total_breaks_v)/2,
                label = paste0('cutoff=',total_breaks_v[first_top]," breaks"),hjust=1))+
  geom_hline(yintercept = first_top,linetype = "dashed") +
    geom_text(aes(total_breaks_v[first_top]-nnn*0.01,first_top,
                label = paste0('cutoff=',first_top," genes"),vjust=1,hjust=1))+
  scale_x_continuous(expand = c(0,NA)) 
#  scale_y_continuous(expand = c(0.5,NA)) 

```

```{r fig2, fig.cap="breaks vs gene rank", out.height='40%'}
ggplot(data=break_rank_gene,aes(x=x,y=y)) +
  geom_line() +
  theme_classic() +
  scale_x_log10() +
  geom_vline(xintercept = first_top,linetype = "dashed") +
  geom_text(aes(first_top,max(total_breaks_v)/2,
                label = paste0('cutoff=',first_top,' genes'),hjust=1)) +
    labs(x="Gene rank (log10)",
       y = "Sonic Breaks")
```


\newpage
## Longitudinal Observation

```{r long_obs_tbl, eval=FALSE, include=FALSE, results='asis'}
kable(
    head(df3.2, n = 50), format = "latex", booktabs = TRUE, 
    caption = paste(
      "Table of top", nrow(head(df3.2, n = 50)), 
      "genes identified by longitudinal observations of patients samples."), 
    align = "c"
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("hold_position","scale_down","striped"),
    font_size = 9
  )
```


```{r long_obs_tbl2,results='asis'}
kbl(df3.2, longtable = T, booktabs = T, caption = paste0(first_top," Genes identified by longitudinal observations.")) %>%
kable_styling(latex_options = c("repeat_header","striped"),font_size = 5.5)

```





