---
params: 
    pTRIAL: "XXXA"
title: 'Longitudinal `r params$pTRIAL`'
#author: "Adrian Cantu (Adapted from Christopher Nobles)"
date: "`r format(Sys.time(), '%Y %B %d')`"
output: 
  pdf_document:
#    latex_engine: lualatex
    toc: false
#    keep_md: true
#    keep_tex: true
    number_sections: false
    toc_depth: 2
    fig_caption: true
    df_print: default
    highlight: espresso
header-includes: 
  - \usepackage{float,indentfirst,booktabs,longtable,array,multirow,pdflscape,tabu}
  - \usepackage[normalem]{ulem}
  - \setlength{\defaultaddspace}{0em}
  - \setlength{\parindent}{2em}
fontsize: 11pt
geometry: margin=0.75in
knit: (
  function(inputFile, encoding) { 
  
    ptrial <- 'CLL'
  
    rmarkdown::render( 
      input       = inputFile, 
      encoding    = encoding, 
      params      = list(pTRIAL = ptrial),      
      output_file = paste(ptrial,'_logitudinal_', Sys.Date(), '.pdf', sep='')) })
---

```{r setup, include=FALSE}
library(tidyverse)
library(kableExtra)

if(!exists('workingDir')) {workingDir <- "/home/ubuntu/data/CART19/CART19_from_git2"}
if(!exists('utilsDir')) {utilsDir <- "/home/ubuntu/data/CART19/CART19_from_git2/utils"}

knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center",
  comment = "",
  echo = FALSE,
  warning = FALSE,
  error = TRUE,
  message = FALSE,
  cache = FALSE,
  dpi = 300,
  dev = "png",
  results = "asis",
  fig.pos = "H",
  fig.width = 7,
  dev.args = list(type = "cairo-png")
)
TRIAL <- params$pTRIAL
```


```{r}
# Import HGNC reference data for annotation and consistency ----
hgnc_complete <- data.table::fread(
  paste0("zcat ", file.path(utilsDir, "hgnc_complete_set.180207.txt.gz")),
  sep = "\t", header = TRUE, 
  select = c(
    "HGNC ID", "Approved Symbol", "Approved Name", "Locus Group", 
    "Locus Type", "Synonyms", "Previous Symbols", "Entrez Gene ID", 
    "Ensembl Gene ID", "RefSeq (supplied by NCBI)", 
    "UniProt ID (supplied by UniProt)"
  ),
  data.table = FALSE
)

names(hgnc_complete) <- c(
  "hgnc_id", "symbol", "name", "locus_group", "locus_type", "alias_symbol", 
  "prev_symbol", "entrez_id", "ensembl_gene_id", "refseq_accession", 
  "uniprot_ids"
)

hgnc_complete <- dplyr::filter(hgnc_complete, !grepl("withdrawn", symbol)) %>%
  dplyr::mutate(
    kegg_id = paste0("hsa:", entrez_id),
    entrez_id = paste0(entrez_id, ":EZID"),
    alias_symbol = gsub(", ", "|", alias_symbol),
    prev_symbol = gsub(", ", "|", prev_symbol),
    extended_alias = paste0(
      alias_symbol, "|", prev_symbol, "|", ensembl_gene_id, "|", 
      refseq_accession, "|", uniprot_ids))

```

```{r}
# redoo relevant parts of gene impact
# gene_impact %>% select(TP_TDN_max_span,max_span,max_time,TP_TDN_max_time)
# df3 %>% select(TP_TDN_max_span,max_span,TP_num_patients,TDN_num_patients,TP_TDN_num_patients)
# stats_tp_tdn %>% filter(TP_TDN_first_time!=TP_TDN_last_time)
#df3
```


```{r}

gene_impact_tmp <- readRDS( # TODO, not sure where this comes from
  file.path(workingDir,paste0('ONLY_',TRIAL),
            paste0("only_",TRIAL,"_cart19_gene_impact.rds"))) 


gene_impact <- gene_impact_tmp %>%
  dplyr::select(
    -ort_fisher_test, -ort_fisher_test_CR, -ort_fisher_test_NR
  ) %>%
  #dplyr::filter(!gene_name %in% excluded_gene_names) %>%
  dplyr::mutate(gene_name = spraphal::alias_arbiter(
    IDs = gene_name,
    RefIDs = hgnc_complete$symbol,
    aliasIDs = hgnc_complete$extended_alias,
    sep = "|", remove_absent_IDs = NULL, quiet = TRUE))

df3 <- gene_impact %>%
  dplyr::filter(
    TP_TDN_num_patients >= 3,
    TP_TDN_num_sites >= 3,
    TP_TDN_sum_abund >= 10,
#    TP_TDN_max_span>5
  )

df3.2 <- df3 %>%
  dplyr::arrange(
    desc(TP_TDN_time_top3avg),desc(TP_TDN_max_span), desc(max_time), desc(TP_TDN_long_count), desc(TP_TDN_num_sites),
    desc(TP_TDN_num_patients), On_Onco_List
  ) %>%
  dplyr::select(
    gene_name, TP_TDN_time_top3avg,TP_TDN_max_span, max_time, TP_TDN_long_count, TP_TDN_num_patients, TP_TDN_num_sites,
    TP_TDN_peak_abund,TP_TDN_peak_relAbund ,On_Onco_List
  ) %>%
  dplyr::mutate(TP_TDN_peak_relAbund=paste0(as.character(round(TP_TDN_peak_relAbund,digits = 2)),'%'),
      TP_TDN_time_top3avg=round(TP_TDN_time_top3avg,digits = 2) ) %>% 
  dplyr::rename(
    "Gene" = gene_name, "Num. Timepoints" = TP_TDN_long_count, "Time Span" = TP_TDN_max_span,
    "Longest Time" = max_time, "Patients" = TP_TDN_num_patients,
    "Sites" = TP_TDN_num_sites, "Max Abund." = TP_TDN_peak_abund,
    "Max RelAbund" = TP_TDN_peak_relAbund,
    "Onco-Related" = On_Onco_List, "t3avg"=TP_TDN_time_top3avg
  )

```

Insertion sites that are close to Genes (less than 5,000bp) are grouped and labeled with that
gene name. Longitudinal gene of interest are those that have insertion sites in at least 3 different positions in at least 3 patients and a total of at least 10 independent sonic breaks. 
The table has the following columns:

* **gene**: Gene name according to the HUGO Gene nomenclature committee (version from 2018-02-07)
* **Time Span**: Maximum time between the first and last sample that have the same insertion in this gene and originate from the same patient.
* **t3avg**: average of last time point for the 3 latest insertion sites.
* **Longest time**: Last time point(in days) where insertion are observed.
* **Num. Timepoints**: Number of times points the most common insertion site is observed.
* **Patients**: Number of patient with at least one insertion near this gene.
* **Sites**: Number of distinct insertion positions.
* **Max Abund**: Maximum sonic abundance observed for an insertion site near this gene.
* **Max RelAbund**: Maximum relative abundance observed for an insertion site near this gene.
* **Onco-Related**: Wheter the gene is in the cancer related list

\newpage
## Longitudinal Observation

```{r long_obs_tbl, eval=FALSE, include=FALSE, results='asis'}
kable(
    head(df3.2, n = 50), format = "latex", booktabs = TRUE, 
    caption = paste(
      "Table of top", nrow(head(df3.2, n = 50)), 
      "genes identified by longitudinal observations of patients samples."), 
    align = "c"
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("hold_position","scale_down","striped"),
    font_size = 9
  )
```


```{r long_obs_tbl2,results='asis'}
kbl(df3.2, longtable = T, booktabs = T, caption = "Genes identified by longitudinal observations .") %>%
#add_header_above(c(" ", "Group 1" = 5, "Group 2" = 6)) %>%
kable_styling(latex_options = c("repeat_header","striped"),font_size = 5.5)

```



