---
params: 
  pTRIAL: "ALL"
  tPART: 'NoNE'
#title: "r params$tPART Patients"
title: |
  | Scan Stat clusters
  | `r params$tPART` Patients and
  | Response Groups (CR/PRtd & PR/NR & NA)
#author: "Adrian Cantu (Adapted from Christopher Nobles)"
date: "`r format(Sys.time(), '%Y %B %d')`"
output: 
  pdf_document:
#    latex_engine: lualatex
    toc: false
#    keep_md: true
#    keep_tex: true
    number_sections: false
    toc_depth: 2
    fig_caption: true
    df_print: default
    highlight: espresso
header-includes: 
  - \usepackage{float,makecell,indentfirst,booktabs,longtable,array,multirow,pdflscape,tabu}
  - \usepackage[normalem]{ulem}
  - \setlength{\defaultaddspace}{0em}
  - \setlength{\parindent}{2em}
fontsize: 11pt
geometry: margin=0.75in
knit: (
  function(inputFile, encoding) { 
  
    ptrial <- 'ALL'
  
    rmarkdown::render( 
      input       = inputFile, 
      encoding    = encoding, 
      params      = list(pTRIAL=ptrial,tPART=ifelse(ptrial=='ALL','ALL',ifelse(ptrial=='CLL','CLL',ifelse(ptrial=='CALL','ALL and CLL','none')))),      
      output_file = paste(ptrial,'_Scan_stats_', Sys.Date(), '.pdf', sep='')) })
---

```{r setup, include=FALSE}
library(tidyverse)
library(kableExtra)
library(gintools)
library(GenomicRanges)


if(!exists('workingDir')) {workingDir <- "/home/ubuntu/data/CART19/CART19_from_git2"}
if(!exists('utilsDir')) {utilsDir <- "/home/ubuntu/data/CART19/CART19_from_git2/utils"}



knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center",
  comment = "",
  echo = FALSE,
  warning = FALSE,
  error = TRUE,
  message = FALSE,
  cache = FALSE,
  dpi = 300,
  dev = "png",
  results = "asis",
  fig.pos = "H",
  fig.width = 7,
  dev.args = list(type = "cairo-png")
)
TRIAL <- params$pTRIAL
```

```{r}


annotate_scan_clusters <- function(scanned_ranges, tdn_sites, 
                                   timepoint_sites, refGenes){
  scanned_ranges$n_sites_tdn <- scan_sites_count(
    scanned_ranges, tdn_sites)
  scanned_ranges$n_sites_tp <- scan_sites_count(
    scanned_ranges, timepoint_sites)
  scanned_ranges$sum_abund_tdn <- scan_sites_abundance(
    scanned_ranges, tdn_sites)
  scanned_ranges$sum_abund_tp <- scan_sites_abundance(
    scanned_ranges, timepoint_sites)
  scanned_ranges$n_patients_tdn <- scan_patients(
    scanned_ranges, tdn_sites)
  scanned_ranges$n_patients_tp <- scan_patients(
    scanned_ranges, timepoint_sites)
  scanned_ranges$genes_in_cluster <- scan_genes(
    scanned_ranges, refGenes)
  scanned_ranges$in_gene_ort_tdn <- scan_orientation(
    scanned_ranges, tdn_sites)
  scanned_ranges$in_gene_ort_tp <- scan_orientation(
    scanned_ranges, timepoint_sites)
  scanned_ranges$ort_fisher_test <- scan_fisher_test_ort(
    scanned_ranges$in_gene_ort_tdn, scanned_ranges$in_gene_ort_tp)
  scanned_ranges
}

scan_sites_count <- function(clus_gr, sites){
  sapply(1:length(clus_gr), function(i){
    clus <- clus_gr[i]
    hits <- subjectHits(findOverlaps(clus, sites))
    length(unique(hits))
  })
}

scan_sites_abundance <- function(clus_gr, sites){
  sapply(1:length(clus_gr), function(i){
    clus <- clus_gr[i]
    hits <- subjectHits(findOverlaps(clus, sites))
    sum(sites[hits]$estAbund)
  })
}
scan_patients <- function(clus_gr, sites){
  sapply(1:length(clus_gr), function(i){
    clus <- clus_gr[i]
    hits <- subjectHits(findOverlaps(clus, sites))
    length(unique(sites[hits]$patient))
  })
}

scan_genes <- function(clus_gr, refGenes){
  sapply(1:length(clus_gr), function(i){
    clus <- clus_gr[i]
    hits <- subjectHits(findOverlaps(clus, refGenes))
    paste(unique(refGenes[hits]$name2), collapse = ", ")
  })
}

scan_orientation <- function(clus_gr, sites){
  sapply(1:length(clus_gr), function(i){
    clus <- clus_gr[i]
    site_hits <- findOverlaps(clus, sites)
    site_ort <- strand(sites[subjectHits(site_hits)])
    gene_ort <- sites[subjectHits(site_hits)]$in_geneOrt
    if(length(sites[subjectHits(site_hits)]) > 0){
      df <- data.frame(
        "posid" = generate_posid(sites[subjectHits(site_hits)]),
        "site_ort" = as.character(site_ort),
        "gene_ort" = as.character(gene_ort),
        "patient" = sites[subjectHits(site_hits)]$patient,
        stringsAsFactors = FALSE
      )
      df <- distinct(df)
      df$same_orientation <- df$site_ort == df$gene_ort
      score <- paste0(
        "T", length(grep("TRUE", df$same_orientation)), ":",
        "F", length(grep("FALSE", df$same_orientation)), ":",
        "N", length(grep("TRUE", is.na(df$same_orientation)))
      )
    }else{
      score <- "T0:F0:N0"
    }
    score
  })
}

scan_fisher_test_ort <- function(score1, score2){
  sapply(1:length(score1), function(i){
    grp1 <- unlist(strsplit(score1[i], ":"))
    grp2 <- unlist(strsplit(score2[i], ":"))
    x <- matrix(c(
      as.integer(substr(grp1[1], 2, 3)),
      as.integer(substr(grp1[2], 2, 3)),
      as.integer(substr(grp2[1], 2, 3)),
      as.integer(substr(grp2[2], 2, 3))),
      ncol = 2
    )
    fisher.test(x)$p.value
  })
}


```

```{r}

refGenes <- readRDS(file.path('..','utils', "hg38.refSeq.rds"))
cond_uniq_sites <- readRDS(file.path(workingDir,paste0('ONLY_',TRIAL),paste0("only_",TRIAL,"_condensed_intsites.rds")))
tdn_sites <- cond_uniq_sites[cond_uniq_sites$timepoint == "d0"]
timepoint_sites <- cond_uniq_sites[cond_uniq_sites$timepoint != "d0"]
```



```{r}
clust_name <- file.path(workingDir,'ONLY_ALL','cluster_reports',paste0("only_",TRIAL,"_long_clusters.rds"))

is_cache <- FALSE
if(file.exists(clust_name)) {
 if(as.Date(file.info(clust_name)$ctime) == Sys.Date()) {
  is_cache <- TRUE 
 }
}

if( !is_cache ) {
  scan_sites <- gintools:::scan_format(
  tdn_sites, timepoint_sites, grouping = "patient") # use "patient" for the same insertion in multiple timepoints on the same patient to be counted as one, use "specimen" to be counted multiple times
  
scanned_clus <- geneRxCluster::gRxCluster(
  object = scan_sites$chr, 
  starts = scan_sites$pos, 
  group = scan_sites$grp,
  kvals = c(10L:50L),
  nperm = 100L,
  cutpt.tail.expr = critVal.target(k, n, target = 7.5, posdiff = x),
  cutpt.filter.expr = apply(x, 2, quantile, probs = 0.35, na.rm = TRUE))

  scanned_clus <- annotate_scan_clusters(
    scanned_clus, tdn_sites, timepoint_sites, refGenes)

  scanned_clus <- scanned_clus[ # make sure you have at least one site per cluster. I'm don't know how this code would be necessary
    scanned_clus$n_patients_tdn > 1 | scanned_clus$n_patients_tp > 1]
  saveRDS(scanned_clus,file = clust_name)
}else{
  scanned_clus <- readRDS(file = clust_name)
}



```

This report list the clusters of insertion sites identified using Scan Statistics ([**Berry *et al.* 2014**](https://academic.oup.com/bioinformatics/article/30/11/1493/282534)) as significantly different between Pre- and Post-infusion samples.

The table has the following columns:

* **Cluster**: location on the genome
* **width**: width of the cluster
* **num PreInf sites**: number of sites in this interval in the pre-infusion (time 0) samples
* **num PostInf sites**: number of sites in this interval in the post-infusion (time 0) samples
* **target.min**: score of the cluster
* **genes**: genes that intersect the cluster



```{r needname,results='asis'}
#linebreak(genes_in_cluster,align = 'l',linebreaker = ',')) Clusters identified by longitudinal observations.
scanned_clus %>% 
  as.data.frame(row.names = NULL) %>% 
  mutate(cluster=paste0(seqnames,' [',gt23::ppNum(start),'-',gt23::ppNum(end),']')) %>% 
  select(c(cluster,width,n_sites_tdn,n_sites_tp,target.min,genes_in_cluster)) %>% 
  mutate(width=gt23::ppNum(width)) %>% 
  mutate(genes_in_cluster=linebreak(genes_in_cluster,align = 'l',linebreaker = ',')) %>% 
  dplyr::rename('num PreInf sites'=n_sites_tdn) %>% 
  dplyr::rename('num PostInf sites'=n_sites_tp) %>% 
  dplyr::rename('genes'=genes_in_cluster) %>% 
  kbl(booktabs = T,longtable = T, escape=FALSE,caption ="Clusters identified by longitudinal observations",linesep = '\\midrule') %>%
  kable_styling(latex_options = c("repeat_header"),font_size = 6)
```

